\documentclass[12pt, leqno]{article}
\input{common}

\begin{document} \hdr{Final}{2016-05-19}

For the final, you are allowed to use texts, papers, or other
references (with citation).  You should not ask help from any other
person, whether inside or outside the class.  You should not worry if
you do not get all the answers; this is a sign that the test is doing
a proper job!  We reserve the right to ask follow-up questions in
person (e.g.~to determine whether it makes sense to assign partial
credit).  You may ask us to clarify ambiguities or perceived errors in
the prompt, but please do not ask for hints.

In addition to a PDF document detailing any derivation work,
you should submit a MATLAB file with your solutions, following
the format of the \verb|final_codes.m| file provided on the web page.
{\em Include tests}.  Clearly untested code (e.g.~code that fails to
execute) will receive at most half credit.

\begin{enumerate}
\item
  Consider the matrix iteration
  \[
    X_{k+1} = X_k (2I-A X_k)
  \]
  where $X_k, A \in \bbR^n$ and $A$ is nonsingular.  There is a unique
  fixed point such that $X$ is nonsingular; characterize this
  fixed point and prove quadratic convergence from good initial
  guesses.

\item
  Give an example of $A \in \bbR^{2 \times 2}$ for which power iteration fails
  to converge for almost all random starting vectors.

\item
  Write a routine to minimize $\|A-xx^T\|_F^2$ for symmetric
  $A \in \bbR^{n \times n}$.
  Do {\em not} assume $A$ is positive definite.

\item
  Write a routine to compute $Q^T A Q = \tilde{H}$ where
  $Q^T c = \alpha e_1$ and $\tilde{H}_{ij} = 0$ for $i > j+2$
  (Hessenberg plus one more subdiagonal).

\item
  Suppose the Jacobian of $f : \bbR^n \rightarrow \bbR^n$ is
  tridiagonal.  Write a routine to estimate the Jacobian by
  finite differences in four function evaluations.

\item
  Write a Newton iteration to solve the system
  \[
    x_{k-1} - (3-x_k/2) x_k + 2 x_{k+1} = 1,
  \]
  for $x_0 = x_{N+1} = 0$.  Use the initial guess $x_k = -1$ for $k=1,
  \ldots, N$.  Your code should run in $O(N)$ time per iteration.

\item
  Suppose $A \in \bbR^{n \times n}$ is SPD and let
  \[
    \phi(x) = \frac{1}{2} x^T A x - x^T b + z^4, \quad
    z = \sum_i x_i.
  \]
  Find the {\em global} minimum of $\phi$ using two solves with $A$.

\item
  The IRLS method for robust linear regression based on minimizing the
  Huber loss function is a fixed-point iteration starting from
  $x^0 = A^\dagger b$:
  \[
    \mbox{minimize } \|\operatorname{diag}(w(r^{k-1})) r^k\|
  \]
  where $r^k = Ax^k-b$ is the residual at step $k$ and $w(r^{k-1})$
  denotes elementwise application of the function
  \[
    w(e) = \begin{cases} 1, & |e| \leq s \\ s/|e|, & |e| > s. \end{cases}
  \]
  Modify this IRLS approach to minimize the Huber loss subject to the
  linear constraint $C^T x = d$ where $C \in \bbR^{n \times p}$ has
  full column rank.  Use the scale factor
  $s = 2 \operatorname{median}(|r_i^0|)$.
    
\end{enumerate}

\end{document}
