\documentclass[12pt, leqno]{article}
\input{common}

\begin{document} \hdr{Midterm}{2016-03-11}

For the midterm, you are allowed to use texts, papers, or other
references (with citation).  You should not ask help from any other
person, whether inside or outside the class.  You should not worry if
you do not get all the answers; this is a sign that the test is doing
a proper job!  We reserve the right to ask follow-up questions in
person (e.g.~to determine whether it makes sense to assign partial
credit).  You may ask us to clarify ambiguities or perceived errors in
the prompt, but please do not ask for hints.

In addition to a PDF document detailing any derivation work,
you should submit a MATLAB file with your solutions, following
the format of the \verb|mt_codes.m| file provided on the web page.

\begin{enumerate}
  
\item Given $A = QR$, solve $Ax = b$ in $O(n^2)$ time.
    
\item Suppose $A = QR$ is an economy QR with $A \in \bbR^{m \times n}$
  and $m \gg n$.
  Minimize $\|Ax-b\|^2 + \lambda^2 \|x\|^2$ with
  $O(mn) + O(n^3)$ additional work.
    
\item Suppose $A$ is symmetric and positive definite and $A = LU$ is
  an ordinary LU factorization.  Show how to get the Cholesky
  factorization $A = R^T R$ in $O(n^2)$ additional time.

\item
  Write a function to compute $\sqrt{x+1}-\sqrt{x-1}$ for $x > 1$.
  You should retain accuracy in floating point for $x \gg 1$.

\item
  Write an $O(n)$ time code equivalent to \verb|x = triu(u*v')\b;|

\item
  The matrix $A \in \bbR^{n \times n}$ is upper triangular except for
  a nonzero $\alpha$ in position $(i,j)$ (with $i > j$).  Given an efficient
  ($O(n^2 k)$ time) algorithm to solve $Ax = b$.

\item
  Compute the $R$ factor in the $QR$
  decomposition of $B$ in $O(n^2)$ time for $A$ of the form
  \[
  A = \begin{bmatrix}
    \alpha_1 \\
    & \alpha_2  \\
    & & \ddots \\
    & & & \alpha_{n-1} &  \\
    \beta_1 & \beta_2 & \ldots & \beta_{n-1} & \alpha_n
  \end{bmatrix}.
  \]

\item
  Suppose $A \in \bbR^{m \times n}$ has many repeated rows.  In MATLAB
  notation, $A = W(p,:)$ --- that is, row $i$ of $A$ is the same as
  row $p_i$ of $W \in \bbR^{k \times n}$.  Write a code to minimize
  $\|Ax-b\|^2$ in $O(m) + O(k n^2)$ time, assuming $A$ full rank.
  
\end{enumerate}

\end{document}
