\documentclass[12pt, leqno]{article}
\input{common}

\begin{document} \hdr{Midterm Course Evaluation}{2016-03-25}

We are at the half-way point in the semester, which means you've had
time to get used to the course, but there is still time to potentially
fix things that aren't working.  I would appreciate constructive
feedback on several points related to the course design.  While your
feedback at this point is not anonymous, you will receive credit if
you complete this evaluation, whether you approve of how the course is
going or not (and I won't hold it against you if you don't like the
course --- just tell me how to make it better!).

\begin{enumerate}
\item {\bf Background:} Do you feel comfortable with the level of the
  class relative to your background?  Are there background topics that
  you wish we would spend more or less time discussing in lecture?
\item {\bf Notes:} Do you get most of your information from the course
  notes, the course text, lecture, or other sources?  Do you have
  concrete suggestions for ways to improve the course notes?
\item {\bf Text:} The course notes parallel the textbook in many ways, but are
  deliberately distinct.  Have you found the textbook useful as a
  resource?  If you have looked at other books beyond the required
  text, which ones have you found most helpful?
\item {\bf Homework:} Roughly how much time do you spend per week on homework?
  Do you feel like the homework does a good job of reinforcing
  concepts taught in class and in the notes?  Are there things that
  you wish were different about the homework?
\item {\bf Problems:} Have you looked at the non-homework ``problems
  to ponder?''  Have you found them interesting?  Helpful for
  studying?
\item {\bf Lecture:} Do you have any constructive comments on my
  (Bindel's) lecture style?  Are there things that I could do (or not
  do) that you feel would make lecture more useful to you?
\item {\bf Policies:} Do you feel the course policies are clear and
  fair?  If I teach the course from the same syllabus again, are there
  aspects of the course policies that I should consider changing?
\item {\bf Other:} What else should I know about how the class is
  going?
\end{enumerate}

\end{document}
